---
title:  'Tutorial 11-12: Teaching case E-Mobility'
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This file is part of the lecture Business Intelligence & Analytics (EESYS-BIA-M), Information Systems and Energy Efficient Systems, University of Bamberg.


##Exercise 1
```{r}
# 1. Load the Data.
trips <- read.csv2(file="BIA_GPS_TIME_SERIES_1.csv", header=TRUE)
```


##Exercise 2
```{r}
# 2. How many rows and columns are there?
dim(trips)
```


##Exercise 3
```{r}
# 3. What are the attributes and objects?
names(trips)
unique(trips$ID_TERMINAL)
```


##Exercise 4
```{r}
# 4. What are the measurement scales for the attributes?
summary(trips)
# By reasoning:
# ID_Terminal: Nominal Scale (Name of the Car)
# Timestamp: Interval Scale (no absolute 0)
# Latitude: Interval Scale (0 artificially chosen, "0" latitude is not "no" latitude)
# Longitude: Interval Scale (0 artificially chosen)
# Speed: Ratio Scale (0 km/h is absolute 0)
# Heading: Interval Scale (0 artificially chosen)
# Deltapos: Ratio Scale (0 m is absolute 0)
# Deltatime: Ratio Scale (0 s is absolute 0)
# ID_Panelsession: Nominal Scale (Name of the Entry)
# ID_Locationtype: Nominal Scale (Name of the Entry)
```


##Exercise 5
```{r}
# 5. How many objects are there?
length(unique(trips$ID_TERMINAL))
```


##Exercise 6
```{r}
# 6. Observations during what time period are covered (exact dates)?
print(paste("From ",trips[1,]$TIMESTAMP, " to ",trips[nrow(trips),]$TIMESTAMP))
# Assuming that data is ordered by time.
# Alternative approach:
trips[order(as.Date(substr(trips$TIMESTAMP,1,10), format="%Y-%m-%d")),][1,]
trips[order(as.Date(substr(trips$TIMESTAMP,1,10), format="%Y-%m-%d")),][
  nrow(trips),]
```


##Exercise 7
```{r}
# 7. Calculate basic statistics for the attributes where it does make sense.
# Central Tendency Measures (Mean, median, mode)
# Note: Mode is appropriate for all measurement scales, while mean and median
# only for interval and above

mean(trips$LATITUDE)
mean(trips$LONGITUDE)
mean(trips[which(trips$ID_PANELSESSION==1),]$SPEED)
mean(trips[which(trips$ID_PANELSESSION==1),]$DELTAPOS)
mean(trips[which(trips$ID_PANELSESSION==0),]$DELTATIME)
median(trips[which(trips$ID_PANELSESSION==1),]$SPEED)
median(trips[which(trips$ID_PANELSESSION==0),]$DELTATIME)

summary(trips)

# mode for ID_LOCATIONTYPE:
sort(table(trips$ID_LOCATIONTYPE))
names(sort(-table(trips$ID_LOCATIONTYPE)))[1]

# Dispersion Measures (Range, interquartile range, standard deviation)

# range(trips$SPEED) - this "range" is not what we need - it returns a vector
# containing the minimum and maximum of all the given arguments.
# range(trips$LATITUDE) this "range" is not what we need 
# range(trips$LONGITUDE) this "range" is not what we need 
# range(trips[which(trips$ID_PANELSESSION==0),]$DELTATIME) this "range" is not
# what we need  

# Important: Range, interquartile range and standard deviation are only
# applicable to interval and ratio scales
# not applicable to ID_Terminal, ID_Panelsession and ID_Locationtype

max(trips$SPEED)-min(trips$SPEED)
max(trips$LATITUDE)-min(trips$LATITUDE)
max(trips$LONGITUDE)-min(trips$LONGITUDE)

quantile(trips[which(trips$ID_PANELSESSION==0),]$DELTATIME,
  probs=c(0,0.25,0.5,0.75,1))
sd(trips[which(trips$ID_PANELSESSION==1),]$SPEED)

# Graphic Displays (Histogram, density-plot)
hist(trips[which(trips$ID_PANELSESSION==1),]$SPEED)
plot(density(trips[which(trips$ID_PANELSESSION==1),]$SPEED,from=0))
```


##Exercise 8
```{r}
# 8. Are there some variables that do not make sense?
# See next exercises...
```


##Exercise 9
```{r}
# 9. Have a closer look at the panel session for the vehicles. Can you identify
# some problems?
unique(trips$ID_PANELSESSION)

for(i in unique(trips$ID_TERMINAL)){
  print(paste("ID: ",i , " Entries: ", nrow(trips[which(trips$ID_TERMINAL==i),
    ])," Ignitions: ", (nrow(trips[which(trips$ID_PANELSESSION==0&trips$
    ID_TERMINAL==i),]))))
}
```


##Exercise 10
```{r}
# 10. How would you interpret these anomalies?
# Reasoning: e.g. defective device?
```


##Exercise 11
```{r}
# 11. Are there unknown location types?
unique(trips$ID_LOCATIONTYPE)
```


##Exercise 12
```{r}
# 12. Check if the speed and respective road types are reasonable?
hist(trips$SPEED)
hist(trips[which(trips$ID_LOCATIONTYPE==0),]$SPEED)
hist(trips[which(trips$ID_LOCATIONTYPE==1),]$SPEED)
hist(trips[which(trips$ID_LOCATIONTYPE==2),]$SPEED)
```


##Exercise 13
```{r}
# 13. Which columns do you consider relevant for our data mining problem?
# Eliminate irrelevant data.
# We don't use Heading
trips <- trips[,c("ID_TERMINAL","TIMESTAMP","LATITUDE","LONGITUDE","SPEED",
  "DELTAPOS","DELTATIME","ID_PANELSESSION", "ID_LOCATIONTYPE")]
# Alternative:
# trips[,-grep("HEADING",colnames(trips))]
```


##Exercise 14
```{r}
# 14.Using your data quality check (on page 14), fix the data. Do you have some
# ideas how to improve / repair the data? 

# a) Eliminate the vehicles with wrong values for panel session.
trips <- trips[-which(trips$ID_PANELSESSION<0),]
trips <- trips[-which(trips$ID_PANELSESSION>2),]
trips <- trips[-which(trips$ID_TERMINAL==207),]
trips <- trips[-which(trips$ID_TERMINAL==2565),]

# b) What could you do with the data, based on the abovementioned speed
# characteristics?
# Option 1: Delete the whole dataset
# unique(trips[which(trips$SPEED>300),]$ID_TERMINAL)
# trips <- trips[-which(trips$ID_TERMINAL==268),]
# Option 2: Set threshold a)
# trips[which(trips$SPEED>300),"SPEED"] <- 300
# Option 3: Set threshold b)
# maxSpeed <- mean(trips$SPEED)+6*sd(trips$SPEED)
# trips[which(trips$SPEED>maxSpeed),"SPEED"] <- maxSpeed
# Option 4: Ignore
# 100*length(trips[which(trips$SPEED>300),"SPEED"])/nrow(trips)
```


##Exercise 15
```{r}
# 15. How could you identify a trip?
# Reasoning: Ignition - Driving - Shut-Off
```


##Exercise 16
```{r}
# 16. Identify the trips (derive the new attribute ???trips???, generate records).
# You may use the function cumsum().
trips$TRIPNUMBER <- 0
trips[which(trips$ID_PANELSESSION==0),]$TRIPNUMBER <- 1
for(i in unique(trips$ID_TERMINAL)){
  trips[which(trips$ID_TERMINAL==i),]$TRIPNUMBER <- cumsum(trips[which(trips$
  ID_TERMINAL==i),]$TRIPNUMBER)
}
```


##Exercise 17
```{r}
# 17. How many trips are there?
nrow(trips[which(trips$ID_PANELSESSION==2),])
```


##Exercise 18
```{r}
# 18. How many trips are there for each individual vehicle?
for(i in unique(trips$ID_TERMINAL)){
  print(paste("ID: ", i, " Number of trips: ",nrow(trips[which(trips$
  ID_PANELSESSION==2&trips$ID_TERMINAL==i),]),sep=""))
}

# Please save your data!
write.csv2(trips, "BIA_GPS_TIME_SERIES_2.csv")
```


##Exercise 19
```{r}
# Additional Exercise
# 19. Calculate basic statistics for the new attribute trips:
library(plyr)
tripsNew <- NULL
for(i in unique(trips$ID_TERMINAL)){
  subset <- trips[which(trips$ID_TERMINAL==i),]
  tempDist <- ddply(subset, .(TRIPNUMBER), summarize, TRIPDIST = sum(DELTAPOS))
  subset <- merge(subset,tempDist[,c("TRIPNUMBER", "TRIPDIST")],
                  by = c("TRIPNUMBER"), all.x=TRUE)
  tripsNew <- rbind(tripsNew, subset)
}
trips <- tripsNew

#  Central tendency measures (mean, median, mode) – on the histogram and on the density plot
mean(trips$TRIPDIST)
median(trips$TRIPDIST)
# mode not reasonable here

#  Dispersion measures (range, interquartile range, standard deviation)
max(trips$TRIPDIST)-min(trips$TRIPDIST)
quantile(trips$TRIPDIST,probs=c(0,0.25,0.5,0.75,1))
sd(trips$TRIPDIST)

#  Graphic displays (histogram, density-plot)
hist(trips$TRIPDIST)
plot(density(trips$TRIPDIST,from=0))


```


##Exercise 20
```{r}
# 20. Load your GPS time series from the first GPS time series tutorial and
# install and load  the package "fpc". It contains an implementation of the
# DBSCAN algorithm.
trips <- read.csv2("BIA_GPS_TIME_SERIES_2.csv", header=TRUE)
install.packages("fpc")
library(fpc)
```


##Exercise 21
```{r}
# 21. Pick the vehicle with ID=782. Use the function dbscan() to infer frequently
# visited locations of the driver.
# Selection of appropriate values for parameters Eps and MinPts require testing
# and interpretation of the problem at hand. Eps=0.0012 and MinPts=6 are a good
# start, but feel free to try different parameters.
subset <- trips[which(trips$ID_TERMINAL==782&trips$ID_PANELSESSION==0),]
clusters <- dbscan(subset[,c("LONGITUDE","LATITUDE")], 1200, MinPts = 6)
```


##Exercise 22
```{r}
# 22. How many clusters did you find?
max(clusters$cluster)
```


##Exercise 23
```{r}
# 23. Which cluster represents the presumable home location? Indicate home base
# parking in your GPS time series data.
subset$CLUSTER <- clusters$cluster

clusters # Note that cluster "0" is noise. We can see that cluster "1" is most
# frequently visited, therefore:
home <- 1

# Alternative: find the most frequently visited cluster automatically
home <- as.numeric(names(sort(table(subset[which(subset$CLUSTER!=0),]$CLUSTER),
                              decreasing=TRUE))[1])

subset$HOMEBASE <- 0
subset[which(subset$CLUSTER==home),]$HOMEBASE <- 1
```


##Exercise 24
```{r}
# 24. Optional Exercise: Perform clustering for all vehicles.
trips$HOMEBASE <- 0
for(id in unique(trips$ID_TERMINAL)){
  subset <- trips[which(trips$ID_TERMINAL==id&trips$ID_PANELSESSION==0),]
  clusters <- dbscan(subset[,c("LONGITUDE","LATITUDE")], 1200, MinPts = 6)
  subset$CLUSTER <- clusters$cluster
  home <- as.numeric(names(sort(table(subset[which(subset$CLUSTER!=0),]$CLUSTER),
                                decreasing=TRUE))[1])
  subset$HOMEBASE <- 0
  subset[which(subset$CLUSTER==home),]$HOMEBASE <- 1
  trips[which(trips$ID_TERMINAL==id),]$HOMEBASE <- 0
  trips[which(trips$ID_TERMINAL==id&trips$ID_PANELSESSION==0),]$HOMEBASE <-
    subset$HOMEBASE
}

# EV Simulation and Visualisation of GPS Driving Data
write.csv2(trips, "BIA_GPS_TIME_SERIES_3.csv")
```


##Exercise 25
```{r}
# 25. Load GPS time series with home locations from VC and pick the vehicle with
# ID=782. We use only one vehicle first for better understanding of the simulation.

trips <- read.csv2("BIA_GPS_TIME_SERIES_3.csv", header=TRUE)
subset <- trips[which(trips$ID_TERMINAL==782),]
```


##Exercise 26
```{r}
# 26. Look at the times when the car parks at the home location. How much energy
# could be charged at home? (Note again that charging power is 3.6 kW)
chargingPower <- 3.6
subset$CHARGING <- subset$HOMEBASE * subset$DELTATIME/3600 * chargingPower
```


##Exercise 27
```{r}
# 27. Now look at the distances between each two measurement points. How much
# energy is required to reach each consecutive location? (Note again that on
# average the vehicle consumes 13kWh/100km)
eConsumption <- 13/100/1000 # Energy consumption in kWh per meter
subset$CONSUMPTION <- subset$DELTAPOS * eConsumption
```


##Exercise 28
```{r}
# 28. Combine information derived from questions 26 and 27. What is the state of
# charge of the vehicle at each measurement point? (Note again that the vehicle
# has a 20kWh battery) Begin with a full battery.

batterycap <- 20 # Battery capacity in kWh
subset$SOC <- 0
subset$SOC[1] <- batterycap

for(i in 2:nrow(subset)){
  subset$SOC[i] <- subset$SOC[i-1]+subset$CHARGING[i] - subset$CONSUMPTION[i]
  if(subset$SOC[i]>batterycap){
    subset$SOC[i] <- batterycap
  } else if(subset$SOC[i]<0){
    subset$SOC[i] <- 0
  }
}

# Alternative solution: more efficient
subset$DIFFERENCE <- subset$CHARGING - subset$CONSUMPTION
subset$SOC <- 0
subset$SOC[1] <- batterycap
f <- function(x, y) max(min(x + y, batterycap), 0)
subset$SOC <- Reduce(f, subset$DIFFERENCE, batterycap, accumulate = TRUE)[-1]
```


##Exercise 29
```{r}
# 29. Optional Exercise: Calculate the state of charge at each measurement point
# for each individual car in our dataset.

newTrips <- NULL
for(id in unique(trips$ID_TERMINAL)){
  subset <- trips[which(trips$ID_TERMINAL==id),]
  chargingPower <- 3.6
  subset$CHARGING <- subset$HOMEBASE * subset$DELTATIME/3600 * chargingPower
  subset$CONSUMPTION <- subset$DELTAPOS * eConsumption
  subset$DIFFERENCE <- subset$CHARGING - subset$CONSUMPTION
  subset$SOC <- 0
  subset$SOC[1] <- batterycap
  f <- function(x, y) max(min(x + y, batterycap), 0)
  subset$SOC <- Reduce(f, subset$DIFFERENCE, batterycap, accumulate = TRUE)[-1]
  newTrips <- rbind(newTrips, subset)
}
subset <- newTrips[which(newTrips$ID_TERMINAL==782),]

100*nrow(subset[which(subset$ID_PANELSESSION==2&subset$SOC>0),])/
  nrow(subset[which(subset$ID_PANELSESSION==2),])
```


##Exercise 30
```{r}
# 30. Again pick the vehicle with ID=782. What percentage of destinations could
# be reached fully electrically? (We define destination as measurement entries
# with ID_PANELSESSION=2)

100*nrow(subset[which(subset$ID_PANELSESSION==2&subset$SOC>0),])/
  nrow(subset[which(subset$ID_PANELSESSION==2),])
``` 
